---
title: "Introduction to Time Series I"
author: "Gustavo Magaña López"
date: "4/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document gathers some main concepts in time 
series analysis as well as some examples written 
in [R](https://www.r-project.org/). R is a free software environment for statistical computing and graphics. It compiles and runs on a wide variety of UNIX platforms, Windows and MacOS. 

## Some definitions
Let's start by defining **stochastic processes**: A family of _indexed_ random variables $Z(\omega, t)$ where $\omega$ belongs to a **sample space** and $t$ belongs to an **index set**.

* For a given $\omega$, $Z(\omega, t)$ as a function of $t$ is called a sample function or **realisation**.
* The population that consists of all possible realisations is called the **ensemble**.
* A **time series** is a collection of **random variables** indexed in a *ordered set* representing time periods.

## Examples

1. **White noise**

For example, one of the basic time series is **white noise**, which is a time series
generated from uncorrelated variables, which are most of the time *normally distributed*.

```{r whitenoise}
white.noise <-  rnorm(300,0,1) # N(0,1) variates 
plot(white.noise, type = "l")
#lines(w)
#v = filter(w, sides=2, filter=rep(1/3,3)) # moving average
```

This collection of random variables $e_{t}$ have the following properties:

1. $E(e_{t}) = 0$
2. $E(e_{t}e_{s}) = 0$
3. $E(e^{2}_{t}) = \sigma^{2}$

For this particular case, having only a couple hundred observations, the estimations are slightly biased but we can observe nevertheless that these properties are conserved. Here we have descriptive statistics for the previously generated **realisation** of _gaussian white noise_, as well as a visualisation of its distribution courtesy of [ggplot2](https://ggplot2.tidyverse.org/).

```{r descriptive.stats}
mean(white.noise)
var(white.noise)
library(ggplot2)
ggplot(as.data.frame(white.noise), aes(x=white.noise)) +
  geom_histogram(aes(y=..density..), colour="black", bins=18) +
    geom_density(alpha=.2, fill="blue") 
```

```{r echo=FALSE}
# library(ggplot2)
#ex <- data.frame(1:1000, w)
#colnames(ex) <- c("index", "noise")
#p <- ggplot(ex, aes(x="index", "noise")) + geom_line(noise)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
