---
title: "Introduction to Time Series I"
author: "Gustavo Magaña López"
date: "4/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document gathers some main concepts in time 
series analysis as well as some examples written 
in [R](https://www.r-project.org/). R is a free software environment for statistical computing and graphics. It compiles and runs on a wide variety of UNIX platforms, Windows and MacOS. 

## Some definitions
Let's start by defining **stochastic processes**: A family of _indexed_ random variables $Z(\omega, t)$ where $\omega$ belongs to a **sample space** and $t$ belongs to an **index set**.

* For a given $\omega$, $Z(\omega, t)$ as a function of $t$ is called a sample function or **realisation**.
* The population that consists of all possible realisations is called the **ensemble**.
* A **time series** is a collection of **random variables** indexed in a *ordered set* representing time periods.

## Examples

1. **White noise**

For example, one of the basic time series is **white noise**, which is a time series
generated from uncorrelated variables, which are most of the time *normally distributed*.

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
